---
title: "Phyloseq"
author: "Briteal"
date: "2024-04-18"
output: html_document
---

# load required packages
```{r}
library(phyloseq)
library(dplyr)
library(BiMiCo)
library(ggplot2)
```

# load taxa and seqtab.nochim
```{r}
load("RData/taxa.RData")
load("RData/seqtab.nochim.RData")
```

#import metadata 
```{r}
metadata<-read.csv("metadata.csv", header=TRUE, row.names = 1)
```

#create phyloseq object
```{r}
#make sure the seqtab.nochim and taxa objects are loaded
physeq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE), sample_data(metadata), tax_table(taxa))
physeq
```

# transform sample counts
```{r}
# convert from raw to abundance (a percentage)
physeq <- transform_sample_counts(physeq, function(abund) 1*(abund>0))
physeq
```

# remove the sequence itself and repalce with ASV
```{r}
dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq, dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq # to see which reads account for it, so ASV is where we look for seqs
```

# remove mt and chlorplast matches. remove all non bacti seqs (needs dplyr package to run)
```{r}
physeq <- physeq %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order) )
#subset is gonna take everything that is mt out
physeq
```

#remove all non bacti seqs
```{r}
physeq<-rm_nonbac(physeq)
physeq
```

#save physeq objects to load later
```{r}
save(physeq, file="RData/physeq.RData")
```

#load pyseq objects to start here
```{r}
load ("RData/physeq.RData")
```

# plot bar graph based on phylum (addining functionality from ggplot2, so do library, where we add color)
```{r}
plot_bar(physeq, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill")
#identiy means raw numbers 
```

#create a barplot of relative abundance, all to 100, separated based on body site 
```{r}
#convert to relative abundance
physeq_relabund <- transform_sample_counts(physeq, function(x) x / sum(x))

#barplot
plot_bar(physeq_relabund, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill") + facet_wrap(~body.site, scales = "free")
#if you change body site to subjects, you get 2 graphs
```

# plot alpha diversity based on body.site (do the diff body sites have different diversity)
```{r}
plot_richness(physeq, x="body.site", color= "subject", measures=c("Observed", "Simpson", "Shannon"))
#observed = straight up, simpson (less sesntive to rare taxa, more clustered) and shannon (more sentsitive to rare taxa) take into the abundance 
#subject - can be changed to whatever metadata topic you want to color 
```

#plot alpha diversity based on reported.antibiotic.usage (in the metadata file the x shows _, make sure its .)
```{r}
plot_richness(physeq, x="reported.antibiotic.usage", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
#shows that antibiotic usage lowers the diversity 
```

# plot alpha diversity based on subject, you can change color to shape 
```{r}
plot_richness(physeq, x="subject", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
```

# test for normality 
```{r}
#shows in Data: creates the alpha which has 3 variables with those 34 points, if alpha says empty then error
alpha <- estimate_richness(physeq, measures=c("Observed", "Simpson", "Shannon"))
```


#shapiro-wilk
```{r}
#make three objects that store our results, do shapiro test on that alpha-object-column 
observed <- shapiro.test(alpha$Observed)
shannon <- shapiro.test(alpha$Shannon)
simpson <- shapiro.test(alpha$Simpson)

# not normal = p value less than 0.05, if greater then it is normally distributed, shannon is the only one greater

#print
print(observed)
print(shannon)
print(simpson)
```

# create data frames for statistical analyses
```{r}
#extract sample info from physeq object
samples <- sample_data(physeq)

#if sample is a phyloseq sample_data object, convert it to a data frame
if (class(samples) == "sample_data") { 
  samples <- data.frame(sample_data(samples))
  }

# add a column to alpha with sample names
alpha$sample <- rownames(alpha)

#merge alpha diversity data and sample data
alpha <- merge(alpha, samples, by = "sample")
```

#perform stats based on subject
```{r}
#perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ subject, data = alpha)
test_simpson <- wilcox.test(Simpson ~ subject, data = alpha)
test_shannon <- t.test(Shannon ~ subject, data = alpha)

#Printing the results
print(test_observed)
print(test_simpson)
print(test_shannon)
```

#perform stats based on reported.antibiotic.usage
```{r}
#perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ reported.antibiotic.usage, data = alpha)
test_simpson <- wilcox.test(Simpson ~ reported.antibiotic.usage, data = alpha)
test_shannon <- t.test(Shannon ~ reported.antibiotic.usage, data = alpha)

#Printing the results
print(test_observed)
print(test_simpson)
print(test_shannon)
#significant diff in diversity on antibiotic treated vs not shown by data p value
```

#test for body site
```{r}
kruskal.test(Simpson ~ body.site, data=alpha) # you pick the test pased on the shapiro results
pairwise.wilcox.test(alpha$Simpson, alpha$body.site, p.adjust.method="holm") #based on simpson and body site, my sig comparisons, the tongue and gut are sig diff, and left palm are sig dif
kruskal.test(Observed ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Observed, alpha$body.site, p.adjust.method="holm")
shannonanova <- aov(Shannon ~ body.site, data=alpha)
summary(shannonanova)
TukeyHSD(shannonanova)
```